Resonant Scattering in a Uniform Sphere with Large Optical Depth

### 2021-11-10 ms.tex
Finishing shane's edits to discussion and summary

The emissivity for an impulsive source with energy E, source position $\vec{x}_s$, and frequency $\nu_s$ is 
\be
j_{\rm em} & = & \frac{E}{4\pi} \delta^3(\vec{x} - \vec{x}_s) \delta(\nu-\nu_{\rm s})\delta (t) ,
\label{eq:jem2}
\ee % TODO: Phil said to cite an equation from the appendix here?
The resulting equation for $J(r,\sigma,t)$ is
\be
-3 \frac{k\phi}{c} \frac{\partial J}{\partial t} + \nabla^2 J + \left( \frac{k}{\Delta} \right)^2 \frac{\partial^2 J}{\partial \sigma^2}
& = & - \frac{\sqrt{6} kE}{4\pi \Delta^2} \delta^3(\vec{x}) \delta (\sigma - \sigma_s ) \delta (t).
\label{eq:diffusion_eqn}
\ee
We employ an expansion in terms of spherical Bessel functions in $r$ and Fourier transform in time. The expansion for $J(r, \sigma, t)$ is then
\be
\label{eq:jrsigmat_expansion}
J(r, \sigma, t) = \sum_{n=1}^{\infty} \int_{-\infty}^\infty \frac{d\omega}{2\pi} e^{-i\omega t} j_0\left(\kappa_n r\right) J(n, \sigma, \omega),
\ee
with
\be \label{eq:jnsigmaomega}
J(n, \sigma, \omega) = \frac{2\kappa_n^2}{R} \int_0^R dr\ r^2 j_0(\kappa_n r) \int_{-\infty}^\infty dt\ e^{i\omega t} J(r, \sigma, t).
\ee
Here, $\omega$ describes the time-dependence of J. Though it is written as a function of the photon frequency variable $\sigma$, $J(r, \sigma, t)$ is the specific mean intensity $dE/(dA dt d\nu)$ and is a distribution in $\nu$. The Fourier coefficient $J(n, \sigma, \omega)$ has units $dE/(dA d\nu)$. Using Equation \ref{eq:jnsigmaomega}, we obtain
\be \label{eq:diffusion_plugged_in}
 \left( \frac{3k\phi}{c}i\omega  -   \kappa_n^2 \right) J(n,\sigma,\omega)  &+& \left( \frac{k}{\Delta} \right)^2 \frac{\partial^2J(n,\sigma,\omega)}{\partial\sigma^2} = -\frac{2\kappa_n^2}{R} \frac{\sqrt{6}}{4\pi} \frac{kE}{\Delta^2} \frac{1}{4\pi} \delta(\sigma - \sigma_s).
\ee
At $\sigma=\sigma_s$, continuity must be enforced,
\be \label{eq:matching_condition_1}
J(n, \sigma^-, \omega) = J(n, \sigma^+, \omega),
\ee
and the discontinuity in $dJ/d\sigma$ due to the source is
\be \label{eq:matching_condition_2}
\frac{\partial J(n, \sigma^+, \omega)}{\partial \sigma} - \frac{\partial J(n, \sigma^-, \omega)}{\partial \sigma} & = & 
- \frac{\sqrt{6}}{8} n^2 \frac{E}{kR^3}.
\ee
At large values of $\sigma$ the line profile $\phi$ is small and Eq. \ref{eq:diffusion_plugged_in} becomes
\be \label{eq:diffusion_at_large_sigma}
\frac{\partial^2J}{\partial\sigma^2} \approx \frac{\Delta^2\kappa_n^2}{k^2} J,
\ee
which has solutions 
\be
J(n, \sigma, \omega)\ {\sim}\ e^{\pm \Delta \kappa_n \sigma / k}.
\ee
This approximate solution implies a boundary condition at large $|\sigma|$
\be \label{eq:single_j_derivative}
\frac{\partial J}{\partial \sigma} = \mp \frac{\Delta\kappa_n}{k} J,
\ee
where a negative sign is taken for large $+\sigma$ and a positive sign is taken for large $-\sigma$ to chose the finite solution as $|\sigma|\to 0$. Numerical integrations are performed inward toward $\sigma_s$ over several domains: from large $|\sigma|$ to $\sigma_s$, from large $|\sigma|$ to 0, and from 0 to $\sigma_s$, depending on whether $\sigma_s$ is positive or negative. If $\sigma_s=0$, just two integrations are performed inward from large $|\sigma|$ to 0. Initial values for integration are obtained either by setting $J=1$ and $dJ/d\sigma$ from Equation \ref{eq:single_j_derivative} at large $|\sigma|$ or by matching $J$ and $dJ/d\sigma$ at 0. This gives $J$ and $J'$ on either side of $\sigma_s$, where a prime indicates the derivative $\partial/\partial \sigma$. By enforcing the matching conditions Eqs. \ref{eq:matching_condition_1} and \ref{eq:matching_condition_2}, the eigenfunctions $J(n, \sigma, \omega)$ are obtained over the domain of photon frequencies $\sigma$. Since the solutions are linear in the starting conditions, only two integrations with different starting values are necessary.

@@

The time-dependence of the problem is solved for a $J=0$ boundary condition to characterize the distribution of photon escape times. Solving the differential equation requires a semi-analytic approach, utilizing an expansion in both space and photon frequency. By matching boundary conditions on either side of the photon source frequency $x_s$, the flux at the surface of the sphere is found as a function of $t$ and $\nu$, expressed as a sum over spatial and frequency modes $n$ and $m$, respectively. Calculating additional spatial eigenmodes provides a more accurate model of the physics nearer to line center, but convergence is slow due to each eigenmode's weak dependence on $n$. Additional frequency eigenmodes introduce fast-oscillating terms that improve the accuracy of the sum, as their contributions cancel with components of lower-order terms to better represent the true solution. Integrating the solution over time produces a spectrum that is shown to agree well with the steady-state calculations in Section \ref{sec:steadystate}, provided a sufficient number of terms in the sum. Integrating the solution over frequency leads to a distribution of photon escape times, which can be compared directly with Monte Carlo simulations. The sum over eigenmodes produces an escape-time distribution that broadly captures the behavior shown by Monte Carlo data---a rise at early times, transitioning to exponential decay in the tail of the distribution. 
% TODO: Shane suggests the following places too much emphasis on the number of spatial terms?
It is expected that the accuracy of the rate constant for the tail of the distribution is limited by the number of spatial terms included in the sum over eigenmodes, as the Doppler core can trap photons at high optical depths until they diffuse outward in frequency, weighting the distribution toward later times. This physics is not modeled by our solution for two reasons: 1) our calculations ignored the Gaussian component of the Voigt line profile, leaving the Lorentzian piece which is accurate only in the line wing, and 2) knowing the core is not modeled accurately, we do not include a large enough number of spatial eigenmodes in the sum to resolve it. 
%
However, a fitting function dependent on parameters $a$ and $\tau_0$ is found that adequately represents the escape time distribution within these constraints. The cumulative distribution function could be calculated using a rejection method, which would then be used to sample photon escape times directly rather than computing their travel times by the Monte Carlo method.

This leads in to the primary application of this work. Models of the escape of hydrogen gas in exoplanet atmospheres can be constructed with a treatment of resonant scattering of \lya in spherical geometry. The Monte Carlo method can be used for this problem, but requires a solution of the radiative transfer equation in spherical geometry and is limited by its high computational demand for large $\tau_0$ where there are many photon scatterings before escape. We seek to develop a computationally efficient random walk model in spherical geometry. There are several methods that are commonly used to accelerate Monte Carlo radiation transfer calculations, including core skipping methods \citep{1968ApJ...153..783A,2002ApJ...567..922A} and hybrid diffusion methods \citep{2018MNRAS.479.2065S}. Another approach with wide application is modified random walk methods, such as those discussed in \citet{1984JCoPh..54..508F, 2009A&A...497..155M, 2010A&A...520A..70R}; an outgoing photon is randomly sampled on the surface of the outgoing sphere by drawing its properties from distributions in outgoing frequencies, directions, and escape times, based on solutions to the diffusion equation. A method similar to this has been applied by \citet{2006ApJ...645..792T} to Lyman $\alpha$ transfer using the \cite{1990ApJ...350..216N} solution, but this solution of course does not use the correct frequency-dependent boundary condition at the surface of the sphere. Furthermore, to perform a full radiation hydrodynamic simulation with Monte Carlo acceleration, it will be necessary to calculate radiation forces due to \lya transfer on the faces of each grid cell in the simulation. Similar calculations have been done in \citet{1976ApJ...208..286W} in plane-parallel geometry. However, these solutions are limited to optical depths below $2.5 \times 10^3$. For this work, it would be necessary to model line center optical depths of up to 1 million or more. Such considerations will be taken in future work while developing this solution into a 1D hydrodynamic simulation using radiation forces from accelerated Monte Carlo \lya transfer.

@@

We have examined previous solutions to \lya transfer in the limit of large optical depth, noting that the separation of variables and treatment of the boundary condition in \cite{1973MNRAS.162...43H}, \cite{1990ApJ...350..216N}, \cite{2006ApJ...649...14D} and others does not produce a valid solution to the transfer equation. Here, we have derived the solution in spherical geometry with an appropriate treatment of the surface boundary condition as a frequency-dependent quantity, examining the spectral and time dependence of the resulting equations. The key result is that the errors in the previously-cited works have been quantified via a correction term, $H_{\rm bc}$, which resolves an excess in flux at the spectral peak and a deficit in the line wing of the calculated spectrum of \lya radiation as compared with Monte Carlo. The size of $H_{\rm bc}$ diminishes at larger $\tau_0$, following a $(a\tau_0)^{-1/3}$ scaling. Introducing time-dependence to the transfer equation leads to another solution which is solved numerically with an eigenfunction expansion. We demonstrate that it agrees with the steady-state spectrum for $x_s=0$ when integrated over time, though its rate of numerical convergence is slow and requires a sum over many modes to become accurate. This treatment is incomplete for emission away from line center ($x_s\neq0$), however, since any terms that would allow asymmetry in the spectrum are not present in the sum. The time-dependent solution is utilized to create wait-time distributions for photons escaping the sphere of optically-thick hydrogen gas, which can be sampled in lieu of traditional Monte Carlo simulations that would be computationally demanding. We compare the calculations from the time-dependent solution with Monte Carlo for a sample of $\tau_0$, noting general agreement in the resulting escape time distributions. In future work, this will be used to accelerate Monte Carlo \lya transfer at large optical depths with potential applications in radiation hydrodynamic simulations of the atmospheres of exoplanets.

@@

The transfer equation is \citep{1986rpa..book.....R}
\be
\frac{1}{c} \frac{\partial I}{\partial t} + \vec{n} \cdot \grad I & =& - \left( \alpha_{\rm sc} + \alpha_{\rm abs} \right) I + (1-p) j_{\rm sc} + j_{\rm em}.
\label{eq:rteqn}
\ee
The scattering coefficient, or inverse mean free path to scattering, is 
\be
\alpha_{\rm sc} & = & n_{\rm sc}\, \frac{\pi e^2}{m_e c}\, f\, \frac{\mathcal{H}(x,a)}{\sqrt{\pi} \Delta}.
= k \phi   
\ee
The absorption coefficient $\alpha_{\rm abs}$, or inverse mean free path to true absorption, is a sum over number density of the absorber times absorption cross section. Once the incoming photon has promoted the electron to an excited state, the collisional de-excitation probability is $p$, and hence only a fraction $1-p$ of the excitations lead to re-emission of photons. Photon destruction will not be considered, so $p$ is set to 0 in this work.

@@

For small frequency shifts $\nu-\nu^\prime$, the incoming intensity may be expanded as 
\be
I(\vec{x},\vec{n}^\prime,\nu^\prime) & \simeq  &
I(\vec{x},\vec{n}^\prime,\nu) 
+ 
\frac{\partial I(\vec{x},\vec{n}^\prime,\nu)}{\partial \nu }(\nu^\prime-\nu)
+ \frac{1}{2} \frac{\partial^2 I(\vec{x},\vec{n}^\prime,\nu)}{\partial \nu^2} (\nu^\prime-\nu)^2 + ...
\ee 
and the Fokker-Planck expansion of $j_{\rm sc}$ is
\be
j_{\rm sc}(\vec{x},\vec{n},\nu) & \simeq  & 4\pi k \int d\Omega^\prime \int d\nu^\prime R(\vec{n},\nu; \vec{n}^\prime,\nu^\prime) 
\left[ 
I(\vec{x},\vec{n}^\prime,\nu) + \frac{\partial I(\vec{x},\vec{n}^\prime,\nu)}{\partial \nu }(\nu^\prime-\nu)
+ \frac{1}{2} \frac{\partial^2 I(\vec{x},\vec{n}^\prime,\nu)}{\partial \nu^2 }(\nu^\prime-\nu)^2
\right]
\nonumber \\ 
& =& 
k\phi(\nu) \int d\Omega^\prime g 
\left[ 
I(\vec{x},\vec{n}^\prime,\nu) 
- \frac{\partial I(\vec{x},\vec{n}^\prime,\nu)}{\partial \nu }
\langle \Delta \nu \rangle
+ \frac{1}{2} \frac{\partial^2 I(\vec{x},\vec{n}^\prime,\nu)}{\partial \nu^2 }
\langle \Delta \nu^2 \rangle
\right]
\ee 
To perform the angular integrals, the Eddington approximation for the angular dependence is inserted with the following result
\be
j_{\rm sc} & = & k\phi J - k\phi \Delta \langle u_\parallel \rangle \left( \frac{\partial J}{\partial \nu} - \frac{6}{5} \vec{n} \cdot \frac{\partial \vec{H}}{\partial \nu} \right)
+ \frac{1}{2} \Delta^2 k\phi \left[ 
\frac{\partial^2 J}{\partial \nu^2} \left( \frac{7}{5} \langle u_\parallel^2 \rangle + \frac{3}{10} \right)
- \frac{12}{5} \langle u_\parallel^2 \rangle 
\vec{n} \cdot \frac{\partial^2 \vec{H}}{\partial \nu^2} \right]
\nonumber \\ & \simeq & 
k\phi J - k\phi \Delta \langle u_\parallel \rangle  \frac{\partial J}{\partial \nu} 
+ \frac{1}{2} \Delta^2 k\phi \left( \frac{7}{5} \langle u_\parallel^2 \rangle + \frac{3}{10} \right)
\frac{\partial^2 J}{\partial \nu^2} .
\label{eq:jsc}
\ee
The first term in Eq. \ref{eq:jsc}, $k\phi J$, represents re-emission of the photon through de-excitation of the atom. It cancels the $-k\phi J$ term in Eq. \ref{eq:rteqn} that corresponds to excitation of the atom. The terms involving frequency derivatives of $\vec{H}$, if carried through the calculation,end up giving terms smaller than then largest terms by a factor of $1/x^2$, which is small on the line wing. These terms are ignored from here on.

If only scattering is included, the transfer equation becomes
\be
\frac{1}{c} \frac{\partial }{\partial t} \left( J + 3\vec{n} \cdot \vec{H} \right)
+ \vec{n} \cdot \grad \left( J + 3\vec{n} \cdot \vec{H} \right)
& = & - k\phi \Delta \langle u_\parallel \rangle  \frac{\partial J}{\partial \nu} 
+ \frac{1}{2} \Delta^2 k\phi \left( \frac{7}{5} \langle u_\parallel^2 \rangle + \frac{3}{10} \right)
\frac{\partial^2 J}{\partial \nu^2}.
\ee
Integrating over angle and frequency then gives
\be
\frac{1}{c} \frac{\partial J(\vec{x}) }{\partial t} +  \grad \cdot \vec{H}(\vec{x}) & = & \int d\nu
\left( - k\phi \Delta \langle u_\parallel \rangle  \frac{\partial J}{\partial \nu} 
+ \frac{1}{2} \Delta^2 k\phi \left( \frac{7}{5} \langle u_\parallel^2 \rangle + \frac{3}{10} \right)
\frac{\partial^2 J}{\partial \nu^2}
\right) 
\nonumber \\ & =& 
k \int d\nu J \frac{\partial }{\partial \nu} 
\left( \phi \Delta \langle u_\parallel \rangle
+ \frac{\partial }{\partial \nu}\left[ \frac{1}{2} \phi \Delta^2 
\left( \frac{7}{5} \langle u_\parallel^2 \rangle + \frac{3}{10}  \right) \right]
\right) = 0,
\ee
where $\vec{H}(\vec{x})$ is the frequency integrated flux, and $\grad \cdot \vec{H}(\vec{x})  = 0 $ if there are no sources or sinks of radiation. Integration by parts has been used to factor $J$ out, assuming each term goes to zero at infinity. The quantity inside brackets must be constant, and since each term should go to zero at infinity, that constant is zero. Hence the first and second moments of the frequency shift are related by
\be
\phi \Delta \langle u_\parallel \rangle
& = & -  \frac{\partial }{\partial \nu}\left[ \frac{1}{2} \phi \Delta^2 
\left( \frac{7}{5} \langle u_\parallel^2 \rangle + \frac{3}{10}  \right) 
\right].
\ee
As an example, on the damping wing, the line profile can be approximated as 
\be \label{eq:app:line_profile_wing}
\phi \simeq \frac{a}{\pi x^2 \Delta}
\ee
with $\langle u_\parallel \rangle \simeq 1/x$ and $\langle u_\parallel^2 \rangle \simeq 1/2$, and so this identity is satisfied. The redistribution function can then be rewritten
\be
j_{\rm sc} & \simeq & k\phi J + \frac{1}{2} k \Delta^2 \frac{\partial }{\partial \nu} 
\left[ \phi  \left( \frac{7}{5} \langle u_\parallel^2 \rangle + \frac{3}{10}  \right)\frac{\partial J}{\partial \nu}  \right]
\simeq k\phi J + \frac{1}{2} k \Delta^2 \frac{\partial }{\partial \nu} 
\left( \phi \frac{\partial J}{\partial \nu}  \right)
\ee
where the last equality (see e.g. \citealt{1994ApJ...427..603R}) is valid on the damping wing where $\langle u_\parallel^2 \rangle \simeq 1/2$.
The following equations will use the approximations for the damping wing.

>>>>

The emissivity for an impulsive source with energy E, source position $\vec{x}_s$, and frequency $\nu_s$ is 
\be
j_{\rm em} & = & \frac{E}{4\pi} \delta^3(\vec{x} - \vec{x}_s) \delta(\nu-\nu_{\rm s})\delta (t) ,
\label{eq:jem2}
\ee % TODO: Phil said to cite an equation from the appendix here?
The resulting equation for $J(r,\sigma,t)$ is
\be
-3 \frac{k\phi}{c} \frac{\partial J}{\partial t} + \nabla^2 J + \left( \frac{k}{\Delta} \right)^2 \frac{\partial^2 J}{\partial \sigma^2}
& = & - \frac{\sqrt{6} kE}{4\pi \Delta^2} \delta^3(\vec{x}) \delta (\sigma - \sigma_s ) \delta (t).
\label{eq:diffusion_eqn}
\ee
We employ an expansion in terms of spherical Bessel functions in $r$ and Fourier transform in time. The expansion for $J(r, \sigma, t)$ is then
\be
\label{eq:jrsigmat_expansion}
J(r, \sigma, t) = \sum_{n=1}^{\infty} \int_{-\infty}^\infty \frac{d\omega}{2\pi} e^{-i\omega t} j_0\left(\kappa_n r\right) J(n, \sigma, \omega),
\ee
with
\be \label{eq:jnsigmaomega}
J(n, \sigma, \omega) = \frac{2\kappa_n^2}{R} \int_0^R dr\ r^2 j_0(\kappa_n r) \int_{-\infty}^\infty dt\ e^{i\omega t} J(r, \sigma, t).
\ee
Here, $\omega$ describes the time-dependence of $J$. Though it is written as a function of the photon frequency variable $\sigma$, $J(r, \sigma, t)$ is the specific mean intensity $dE/(dA dt d\nu)$ and is a distribution in $\nu$. The Fourier coefficient $J(n, \sigma, \omega)$ has units $dE/(dA d\nu)$. Using Equation \ref{eq:jnsigmaomega}, we obtain
\be \label{eq:diffusion_plugged_in}
 \left( \frac{3k\phi}{c}i\omega  -   \kappa_n^2 \right) J(n,\sigma,\omega)  &+& \left( \frac{k}{\Delta} \right)^2 \frac{\partial^2J(n,\sigma,\omega)}{\partial\sigma^2} = -\frac{2\kappa_n^2}{R} \frac{\sqrt{6}}{4\pi} \frac{kE}{\Delta^2} \frac{1}{4\pi} \delta(\sigma - \sigma_s).
\ee
At $\sigma=\sigma_s$, continuity must be enforced,
\be \label{eq:matching_condition_1}
J(n, \sigma^-, \omega) = J(n, \sigma^+, \omega),
\ee
and the discontinuity in $dJ/d\sigma$ due to the source is
\be \label{eq:matching_condition_2}
\frac{\partial J(n, \sigma^+, \omega)}{\partial \sigma} - \frac{\partial J(n, \sigma^-, \omega)}{\partial \sigma} & = & 
- \frac{\sqrt{6}}{8} n^2 \frac{E}{kR^3}.
\ee
At large values of $\sigma$ the line profile $\phi$ is small and Eq. \ref{eq:diffusion_plugged_in} becomes
\be \label{eq:diffusion_at_large_sigma}
\frac{\partial^2J}{\partial\sigma^2} \approx \frac{\Delta^2\kappa_n^2}{k^2} J,
\ee
which has solutions 
\be
J(n, \sigma, \omega)\ {\sim}\ e^{\pm \Delta \kappa_n \sigma / k}.
\ee
This approximate solution implies a boundary condition at large $|\sigma|$
\be \label{eq:single_j_derivative}
\frac{\partial J}{\partial \sigma} = \mp \frac{\Delta\kappa_n}{k} J,
\ee
where a negative sign is taken for large $+\sigma$ and a positive sign is taken for large $-\sigma$ to chose the finite solution as $|\sigma|\to 0$. Numerical integrations are performed inward toward $\sigma_s$ over several domains: from large $|\sigma|$ to $\sigma_s$, from large $|\sigma|$ to 0, and from 0 to $\sigma_s$, depending on whether $\sigma_s$ is positive or negative. If $\sigma_s=0$, just two integrations are performed inward from large $|\sigma|$ to 0. Initial values for integration are obtained either by setting $J=1$ and $dJ/d\sigma$ from Equation \ref{eq:single_j_derivative} at large $|\sigma|$ or by matching $J$ and $dJ/d\sigma$ at 0. This gives $J$ and $J'$ on either side of $\sigma_s$, where a prime indicates the derivative $\partial/\partial \sigma$. By enforcing the matching conditions Eqs. \ref{eq:matching_condition_1} and \ref{eq:matching_condition_2}, the eigenfunctions $J(n, \sigma, \omega)$ are obtained over the domain of photon frequencies $\sigma$. Since the solutions are linear in the starting conditions, only two integrations with different starting values are necessary.

@@

The time-dependence of the problem is solved for a $J=0$ boundary condition to characterize the distribution of photon escape times. Solving the differential equation requires a semi-analytic approach, utilizing an expansion in both space and photon frequency. By matching boundary conditions on either side of the photon source frequency $x_s$, the flux at the surface of the sphere is found as a function of $t$ and $\nu$, expressed as a sum over spatial and frequency modes $n$ and $m$, respectively. Calculating additional spatial eigenmodes provides a more accurate model of the physics nearer to line center, but convergence is slow due to each eigenmode's weak dependence on $n$. Additional frequency eigenmodes introduce fast-oscillating terms that improve the accuracy of the sum, as their contributions cancel with components of lower-order terms to better represent the true solution. Integrating the solution over time produces a spectrum that is shown to agree well with the steady-state calculations in Section \ref{sec:steadystate}, provided a sufficient number of terms in the sum. Integrating the solution over frequency leads to a distribution of photon escape times, which can be compared directly with Monte Carlo simulations. The sum over eigenmodes produces an escape-time distribution that broadly captures the behavior shown by Monte Carlo data---a rise at early times, transitioning to exponential decay in the tail of the distribution. It is expected that the accuracy of the rate constant for the tail of the distribution is limited by the effect of the Doppler core, which can trap photons at high optical depths until they diffuse outward in frequency, weighting the distribution toward later times. This physics is not modeled by our solution for two reasons: 1) our calculations ignored the Gaussian component of the Voigt line profile, leaving the Lorentzian piece which is accurate only in the line wing, and 2) knowing the core is not modeled accurately, we do not include a large enough number of spatial eigenmodes in the sum to resolve it. However, an approximate fitting function dependent on parameters $a$ and $\tau_0$ is found that adequately represents the escape time distribution within these constraints.

This leads in to the primary application of this work. Models of the escape of hydrogen gas in exoplanet atmospheres can be constructed with a treatment of resonant scattering of \lya in spherical geometry. The Monte Carlo method can be used for this problem, but is limited by its high computational demand for large $\tau_0$ where there are many photon scatterings before escape. We seek to develop a computationally efficient random walk model which requires a solution in spherical geometry. There are several methods that are commonly used to accelerate Monte Carlo radiation transfer calculations, including core skipping methods \citep{1968ApJ...153..783A,2002ApJ...567..922A} and hybrid diffusion methods \citep{2018MNRAS.479.2065S} but we believe that these methods are not sufficient for our application. Another approach with wide application is modified random walk methods, such as those discussed in \citet{1984JCoPh..54..508F, 2009A&A...497..155M, 2010A&A...520A..70R}; an outgoing photon is randomly sampled on the surface of the outgoing sphere by drawing its properties from distributions in outgoing frequencies, directions, and escape times, based on solutions to the diffusion equation. A method similar to this has been applied by \citet{2006ApJ...645..792T} to Lyman $\alpha$ transfer using the \cite{1990ApJ...350..216N} solution, but this solution of course does not use the correct frequency-dependent boundary condition at the surface of the sphere. Furthermore, to perform a full radiation hydrodynamic simulation with Monte Carlo acceleration, it will be necessary to calculate radiation forces due to \lya transfer on the faces of each grid cell in the simulation. Similar calculations have been done in \citet{1976ApJ...208..286W} in plane-parallel geometry. However, these solutions are limited to optical depths below $2.5 \times 10^3$. For this work, it would be necessary to model line center optical depths of up to 1 million or more. We will utilize the solution derived here as the basis for our own novel implementation of the modified random walk method.

@@

We have examined previous solutions to \lya transfer in the limit of large optical depth, noting that the separation of variables and treatment of the boundary condition in \cite{1973MNRAS.162...43H}, \cite{1990ApJ...350..216N}, \cite{2006ApJ...649...14D} and others does not produce a valid solution to the transfer equation. Here, we have derived the solution in spherical geometry with an appropriate treatment of the surface boundary condition, examining the spectral and time dependence of the resulting equations. The key result is that the errors in the previously-cited works have been quantified via a correction term, $H_{\rm bc}$, which resolves an excess in flux at the spectral peak and a deficit in the line wing of the calculated spectrum of \lya radiation as compared with Monte Carlo. The size of $H_{\rm bc}$ is of order unity when the spectral peaks are near the Doppler core, and diminishes at larger $\tau_0$ following a $(a\tau_0)^{-1/3}$ scaling. Introducing time-dependence to the transfer equation leads to another solution which is solved numerically with an eigenfunction expansion. We demonstrate that it agrees with the steady-state spectrum for $x_s=0$ when integrated over time, though its rate of numerical convergence is slow and requires a sum over many modes to become accurate. This treatment is incomplete for emission away from line center ($x_s\neq0$), however, since any terms that would allow asymmetry in the spectrum are not present in the sum. The time-dependent solution is utilized to create wait-time distributions for photons escaping the sphere of optically-thick hydrogen gas, which can be sampled in lieu of traditional Monte Carlo simulations that would be computationally demanding. We compare the calculations from the time-dependent solution with Monte Carlo for a sample of $\tau_0$, noting general agreement in the resulting escape time distributions. In future work, this will be used to accelerate Monte Carlo \lya transfer at large optical depths with potential applications in radiation hydrodynamic simulations of the atmospheres of exoplanets.

@@

The transfer equation is \citep{1986rpa..book.....R}
\be
\frac{1}{c} \frac{\partial I}{\partial t} + \vec{n} \cdot \grad I & =& - \left( \alpha_{\rm sc} + \alpha_{\rm abs} \right) I + (1-p) j_{\rm sc} + j_{\rm em}.
\label{eq:rteqn}
\ee
The scattering coefficient, or inverse mean free path to scattering, is 
\be
\alpha_{\rm sc} & = & n_{\rm sc}\, \frac{\pi e^2}{m_e c}\, f\, \frac{\mathcal{H}(x,a)}{\sqrt{\pi} \Delta}
= k \phi.
\ee
The absorption coefficient $\alpha_{\rm abs}$, or inverse mean free path to true absorption, is a sum over number density of the absorber times absorption cross section. Once the incoming photon has promoted the electron to an excited state, the collisional de-excitation probability is $p$, and hence only a fraction $1-p$ of the excitations lead to re-emission of photons. Photon destruction will not be considered, so $p$ is set to 0 in this work.

@@

For small frequency shifts $\nu-\nu^\prime$, the incoming intensity may be expanded as 
\be
I(\vec{x},\vec{n}^\prime,\nu^\prime) & \simeq  &
I(\vec{x},\vec{n}^\prime,\nu) 
+ 
\frac{\partial I(\vec{x},\vec{n}^\prime,\nu)}{\partial \nu }(\nu^\prime-\nu)
+ \frac{1}{2} \frac{\partial^2 I(\vec{x},\vec{n}^\prime,\nu)}{\partial \nu^2} (\nu^\prime-\nu)^2 + ...
\ee 
and the Fokker-Planck expansion of $j_{\rm sc}$ is
\be
j_{\rm sc}(\vec{x},\vec{n},\nu) & \simeq  & 4\pi k \int d\Omega^\prime \int d\nu^\prime R(\vec{n},\nu; \vec{n}^\prime,\nu^\prime) 
\left[ 
I(\vec{x},\vec{n}^\prime,\nu) + \frac{\partial I(\vec{x},\vec{n}^\prime,\nu)}{\partial \nu }(\nu^\prime-\nu)
+ \frac{1}{2} \frac{\partial^2 I(\vec{x},\vec{n}^\prime,\nu)}{\partial \nu^2 }(\nu^\prime-\nu)^2
\right]
\nonumber \\ 
& =& 
k\phi(\nu) \int d\Omega^\prime g 
\left[ 
I(\vec{x},\vec{n}^\prime,\nu) 
- \frac{\partial I(\vec{x},\vec{n}^\prime,\nu)}{\partial \nu }
\langle \Delta \nu \rangle
+ \frac{1}{2} \frac{\partial^2 I(\vec{x},\vec{n}^\prime,\nu)}{\partial \nu^2 }
\langle \Delta \nu^2 \rangle
\right].
\ee 
To perform the angular integrals, the Eddington approximation for the angular dependence is inserted with the following result
\be
j_{\rm sc} & = & k\phi J - k\phi \Delta \langle u_\parallel \rangle \left( \frac{\partial J}{\partial \nu} - \frac{6}{5} \vec{n} \cdot \frac{\partial \vec{H}}{\partial \nu} \right)
+ \frac{1}{2} \Delta^2 k\phi \left[ 
\frac{\partial^2 J}{\partial \nu^2} \left( \frac{7}{5} \langle u_\parallel^2 \rangle + \frac{3}{10} \right)
- \frac{12}{5} \langle u_\parallel^2 \rangle 
\vec{n} \cdot \frac{\partial^2 \vec{H}}{\partial \nu^2} \right]
\nonumber \\ & \simeq & 
k\phi J - k\phi \Delta \langle u_\parallel \rangle  \frac{\partial J}{\partial \nu} 
+ \frac{1}{2} \Delta^2 k\phi \left( \frac{7}{5} \langle u_\parallel^2 \rangle + \frac{3}{10} \right)
\frac{\partial^2 J}{\partial \nu^2} .
\label{eq:jsc}
\ee
The first term in Eq. \ref{eq:jsc}, $k\phi J$, represents re-emission of the photon through de-excitation of the atom. It cancels the $-k\phi J$ term in Eq. \ref{eq:rteqn} that corresponds to excitation of the atom. The terms involving frequency derivatives of $\vec{H}$, if carried through the calculation,end up giving terms smaller than then largest terms by a factor of $1/x^2$, which is small on the line wing. These terms are ignored from here on.

If only scattering is included, the transfer equation becomes
\be
\frac{1}{c} \frac{\partial }{\partial t} \left( J + 3\vec{n} \cdot \vec{H} \right)
+ \vec{n} \cdot \grad \left( J + 3\vec{n} \cdot \vec{H} \right)
& = & - k\phi \Delta \langle u_\parallel \rangle  \frac{\partial J}{\partial \nu} 
+ \frac{1}{2} \Delta^2 k\phi \left( \frac{7}{5} \langle u_\parallel^2 \rangle + \frac{3}{10} \right)
\frac{\partial^2 J}{\partial \nu^2}.
\ee
Integrating over angle and frequency then gives
\be
\frac{1}{c} \frac{\partial J(\vec{x}) }{\partial t} +  \grad \cdot \vec{H}(\vec{x}) & = & \int d\nu
\left( - k\phi \Delta \langle u_\parallel \rangle  \frac{\partial J}{\partial \nu} 
+ \frac{1}{2} \Delta^2 k\phi \left( \frac{7}{5} \langle u_\parallel^2 \rangle + \frac{3}{10} \right)
\frac{\partial^2 J}{\partial \nu^2}
\right) 
\nonumber \\ & =& 
k \int d\nu J \frac{\partial }{\partial \nu} 
\left( \phi \Delta \langle u_\parallel \rangle
+ \frac{\partial }{\partial \nu}\left[ \frac{1}{2} \phi \Delta^2 
\left( \frac{7}{5} \langle u_\parallel^2 \rangle + \frac{3}{10}  \right) \right]
\right) = 0,
\ee
where $\vec{H}(\vec{x})$ is the frequency integrated flux, and $\grad \cdot \vec{H}(\vec{x})  = 0 $ if there are no sources or sinks of radiation. Integration by parts has been used to factor $J$ out, assuming each term goes to zero at infinity. The quantity inside brackets must be constant, and since each term should go to zero at infinity, that constant is zero. Hence the first and second moments of the frequency shift are related by
\be
\phi \Delta \langle u_\parallel \rangle
& = & -  \frac{\partial }{\partial \nu}\left[ \frac{1}{2} \phi \Delta^2 
\left( \frac{7}{5} \langle u_\parallel^2 \rangle + \frac{3}{10}  \right) 
\right].
\ee
As an example, on the damping wing, the line profile can be approximated as 
\be \label{eq:app:line_profile_wing}
\phi \simeq \frac{a}{\pi x^2 \Delta}
\ee
with $\langle u_\parallel \rangle \simeq 1/x$ and $\langle u_\parallel^2 \rangle \simeq 1/2$, and so this identity is satisfied. The redistribution function can then be rewritten
\be
j_{\rm sc} & \simeq & k\phi J + \frac{1}{2} k \Delta^2 \frac{\partial }{\partial \nu} 
\left[ \phi  \left( \frac{7}{5} \langle u_\parallel^2 \rangle + \frac{3}{10}  \right)\frac{\partial J}{\partial \nu}  \right]
\simeq k\phi J + \frac{1}{2} k \Delta^2 \frac{\partial }{\partial \nu} 
\left( \phi \frac{\partial J}{\partial \nu}  \right),
\ee
where the last equality (see e.g. \citealt{1994ApJ...427..603R}) is valid on the damping wing where $\langle u_\parallel^2 \rangle \simeq 1/2$.
The following equations will use the approximations for the damping wing.
